Poisson approximate likelihood compared to the particle filter

### 2024-09-15 ms.Rnw

>>>>

\section{Extra material excluded in the final version}

%%%%%%%%% 222222222 %%%%%%%%%%

\section*{Methods}

We first replicated the model of WWR in the R package \texttt{pomp} \citep{pomppackagepaper}.
Comparing summary statistics of simulations confirms that both model versions are essentially identical, as described by \citet{hao24}.
We therefore implement PAL using the code from \citet{wwr}, comparing directly with the particle filter implemented using \texttt{pomp}.

This article has focused on likelihood evaluation, but the inferences presented also require the likelihood estimate to be maximized.
WWR used direct numerical maximization, whereas SBH used iterated filtering to maximize the particle filter likelihood estimate implemented using the \texttt{mif2} algorithm in \texttt{pomp}.
We follow the approaches adopted by these two papers.
The optimization method is not directly pertinent to the comparison of the filtering methods, so we do not discuss optimization in further detail here.

\begin{table}[htbp]
    \caption{Maximum likelihood estimation of parameters for the OvOv model of WWR by PAL (taken from WWR), and by iterated particle filtering implemented using {\rm \texttt{mif2}} in {\rm \texttt{pomp}} \citep{pomppackagepaper}. The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be fixed in \cite{wwr}. The iterated particle filtering estimates were obtained by using 3 rounds of refinement, with 100 filtering iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round.}
    \label{tab:mlebywwr}
\begin{center}    
    \begin{tabular}{lccc}  % Left-aligned for the first column, centered for the other two columns
      \toprule
      Parameter      & PAL      & iterated filtering & iterated filtering (stocks)\\
      \midrule
      \(\beta_1\)    & 11.48     & $\Sexpr{round(est_mle_vec["beta1"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta1"],2)}$\\
      \(\beta_2\)    & 0.25      & $\Sexpr{round(est_mle_vec["beta2"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta2"],2)}$\\
      \(\beta_3\)    & 0.35      & $\Sexpr{round(est_mle_vec["beta3"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta3"],2)}$\\
      \(\phi\)       & 0.14      & $\Sexpr{round(est_mle_vec["phi"],2)}$   & $\Sexpr{round(est_mle_vec_sbh["phi"],2)}$\\
      \(\rho\)       & 0.16      & $\Sexpr{round(est_mle_vec["beta11"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta11"],2)}$\\
      \(\sigma^2_q\) & 0.021     & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_q"],3)}$\\
      \(\sigma_\xi\) & 66.89     & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$ & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$\\
      \(S_{10}\)     & 3876549   & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$  & $5094840$\\
      \(I_{10}\)     & 30351     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$  & $2871$\\
      \(R_{10}\)     & 1315221   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$  & $124410$\\
      \(S_{20}\)     & 57139612  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$  & $57385600$\\
      \(I_{20}\)     & 871       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$  & $639$\\
      \(R_{20}\)     & 302852    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$  & $57072$\\
      \(S_{30}\)     & 19573727  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$  & $19697600$\\
      \(I_{30}\)     & 2550      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$  & $174$\\
      \(R_{30}\)     & 131092    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$  & $9578$\\
      \midrule
      \textbf{AIC}  & 13778.08  & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$ & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec_sbh["logLik"]),2)}$\\
      \bottomrule
    \end{tabular}
\end{center}
\end{table}

<<plot2-cond-loglik, echo=FALSE, fig.height=4, fig.width=5.5, out.width="4.5in",fig.align="center", fig.cap='Conditional log-likelihoods computed using PAL (red) and PF (blue) for the rotavirus model and MLE from WWR. We used 50000 particle for both PF and PAL. PF averaged over 36 replications to reduce the Monte Carlo variance, but this step was not necessary for PAL. The main sources of likelihood shortfall of PF are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.'>>=

# Combine the data into a single dataframe for SMC with grouping variable
data_list <- lapply(1:ncores, function(i) {
  data.frame(
    time = time, 
    cond_loglik = pfilter_wwr_cond_lik[[i]], 
    method = "SMC",
    group = i
  )
})

# Combine all the SMC data frames into one
df_smc <- bind_rows(data_list)

# Create the PAL data frame
df_pal <- data.frame(
  time = time, 
  cond_loglik = palfilter_lik_list_wwr$ll_storage, 
  method = "PAL",
  group = 1  # Single group for PAL
)

# Combine both data frames
df <- bind_rows(df_smc, df_pal)

# Plot with ggplot
ggplot(df, aes(x = time, y = cond_loglik, color = method)) +
  geom_line(data = df %>% filter(method == "SMC"), aes(group = group), color = "blue", alpha = 0.7) +
  geom_line(data = df %>% filter(method == "PAL"), aes(group = group), color = "red") +
  theme_minimal() +
  labs(
    title = NULL,
    x = "Time (weeks)",
    y = "Conditional Log-likelihood",
    color = "Method"
  ) +
  theme(legend.position = "bottom")
@

\begin{table}[htbp]
    \caption{Maximum likelihood estimation of parameters for the OvOv model of WWR by PAL (taken from WWR), and by iterated particle filtering implemented using {\rm \texttt{mif2}} in {\rm \texttt{pomp}} \citep{pomppackagepaper}. The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be fixed in \cite{wwr}. The iterated particle filtering estimates were obtained by using 3 rounds of refinement, with 100 filtering iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round.}
    \label{tab:mlebywwr}
\begin{center}    
    \begin{tabular}{lccc}  % Left-aligned for the first column, centered for the other two columns
      \toprule
      Parameter      & PAL      & iterated filtering & iterated filtering (stocks)\\
      \midrule
      \(\beta_1\)    & 11.48     & $\Sexpr{round(est_mle_vec["beta1"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta1"],2)}$\\
      \(\beta_2\)    & 0.25      & $\Sexpr{round(est_mle_vec["beta2"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta2"],2)}$\\
      \(\beta_3\)    & 0.35      & $\Sexpr{round(est_mle_vec["beta3"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta3"],2)}$\\
      \(\phi\)       & 0.14      & $\Sexpr{round(est_mle_vec["phi"],2)}$   & $\Sexpr{round(est_mle_vec_sbh["phi"],2)}$\\
      \(\rho\)       & 0.16      & $\Sexpr{round(est_mle_vec["beta11"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta11"],2)}$\\
      \(\sigma^2_q\) & 0.021     & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_q"],3)}$\\
      \(\sigma_\xi\) & 66.89     & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_xi"],2)}$\\
      \(S_{10}\)     & 3876549   & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$  & $5094840$\\
      \(I_{10}\)     & 30351     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$  & $2871$\\
      \(R_{10}\)     & 1315221   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$  & $124410$\\
      \(S_{20}\)     & 57139612  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$  & $57385600$\\
      \(I_{20}\)     & 871       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$  & $639$\\
      \(R_{20}\)     & 302852    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$  & $57072$\\
      \(S_{30}\)     & 19573727  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$  & $19697600$\\
      \(I_{30}\)     & 2550      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$  & $174$\\
      \(R_{30}\)     & 131092    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$  & $9578$\\
      \midrule
      \textbf{AIC}  & 13778.08  & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$ & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec_sbh["logLik"]),2)}$\\
      \bottomrule
    \end{tabular}
\end{center}
\end{table}

Figure~\ref{fig:plot2-cond-loglik} indicates that the worst conditional log-likelihood values, estimated using the particle filter, arise at the start of the time series.
This could be a problem with data collection, or some subtle issue with the science of the epidemiological system, but a simple possibility is that the initial values of the latent state process might be poorly specified.
For this model, the initial values were fixed based on scientific reasoning, rather than being estimated.
If we instead estimate the initial values, we obtain the results in Table~\ref{tab:mlebywwr}.
